Motor Learning Without Moving

### 2019-03-30 ManuscriptMaterials.Rmd
organized scripts to reflect revisions

These are the planned figures. Figure one will not be included here, and can in fact simply be re-used from the previous manuscript.

1. Setup & Task (goal: visualize the experiment)
    + A: Setup (side view)
    + B: Rotated training (top view)
    + C: No-cursor reaches (top view)
    + D: Localization (top view)
2. Localization
    + A: Exposure, active and passive (goal: small/no difference between active and passive)
    + B: Classic, active and passive (goal: larger difference between active and passive, different generalization?)
    + C: Online localization?
3. Reach Aftereffects (goal: show motor changes are there, are robust and compare to classic)
    + A: Exposure reach aftereffects (initial & later)
    + B: Classic & Exposure reach aftereffects (initial only? or all?)

@@

4. Correlation
    + A: Exposure, active and passive localization predicting reach aftereffects
    + B: Classic, active and passive localization predicting reach aftereffects

## Tables

We expand the table used in the previous manuscript to include this paradigm as well.

1. Task order & trial numbers (goal: clarify task order and show similarity between classic and exposure experiment - so probably after Fig 1)

@@

The scripts doing the statistics and creating figures are in separate R files. They allow many options, but here we will only see / do the ones that we think should go to the manuscript.

@@

```{r require_lme_packages}
required.packages = c('lme4', 'lmerTest', installmissing=FALSE)
installRequire.Packages(required.packages)
```

@@

In the manuscript we'll first show that there indeed are localization shifts after exposure training (with the no-cursor data) and then compare that with the classic. Second are the no-cursor reaches (and the reach aftereffects). A third topic might be if localization shifts can predict no-cursor changes, which we'll call the "correlation" topic. Here we'll also follow this order:

1. Analyses (some general remarks about analyses)
2. Participants (participant demographics)
3. Localization
4. Reach aftereffects
5. Correlations

@@

Since in the localization tasks there is missing data at some points (usually at the extremes) we can't use ANOVAs directly on the data. Imputation is going to be very hard because the missing data is systematically missing. Hence, we are using Linear Mixed Effects models as an alternative, but this presents a new problem. Getting the usual F-values and p-values is not so straightforward. Initially we decided to use the $\chi^2$ approach that we get from the `Anova` function from the `car` package. However, a paper from Steven G. Luke (2017, _Evaluating significance in linear mixed-effects models in R_) suggests using Satterthwaite approximation. So all functions running an LME now take an argument that tells them to use the Satterthwaite method, but the default is chi-squared. It turns out that it doesn't matter much, so it's a good sanity check. For consistency across all the analyses, we set the value of that argument here:

@@

```{r requireChiSquared_packages}
if (LMEmethod=='chi-squared') {
  required.packages = c('nlme', 'car')
  installRequire.Packages(required.packages, installmissing=FALSE)
}
```

### Peak location of generalization

Since participants trained with a single target, we could look at generalization curves. These appear to be shifted more counter clockwise in most cases, so we want to quantify this. But, for most of the curves, and especially the ones that seems to be shifted, we don't have data on the full curve on one side of the peak. We can still bootstrap the __peak__ of Guassian functions fit to the data across many samples of participants for the exposure group and for the delayed localization after classic training, but not for the online localization after classic training. This data is presented in the paper, where we only look at the peak of the localization -- not the width of the generalization curve.

The range of the data doesn't include the peak in many cases, which makes it hard to fit a normal function to the data. In one alternative approach, we assume that the minimal change in the data, occurs far beyond the boundaries of the measured workspace, make the data cumulative and scale back to a range from 0 to 1. This allows fitting a _cumulative_ normal function where the 50% point corresponds to the peak of the generalization curve. This method is more stable, and seems to works fairly well on the datasets where the more common fits also work (but we didn't test online localization from the classic experiment). Code for this approach is available (functions `psychometricGeneralization()`, `pnormMSE()` and `getCDFpeaks()` at the end of the file: `shared.R`), but the results are not presented in the paper or notebook.

@@

There are no scripts for this part, as the code is simple and can just be embedded.

@@

We'll also set a variable that is to be passed to all functions to exclude the underperformance:

@@

@@

@@

## Localization peaks

There was an interaction between group and hand angle on localization shifts above, which suggests different generalization curves in the two groups. This could reflect a difference in the shape (more flat for the classic training group as compared to exposure), but also a difference in the locations of the peaks of the generalization curves.

Here are the confidence intervals for the peaks of the generalization curves for _active_ localization:

The median values are a little over 13 degrees apart, and they _just_ fall outside each others confidence interval. This can explain the interaction between group and hand angle on localization shifts.

@@

Here is a plot, probably Figure 3:

```{r fig2_reachaftereffect, fig.width=7.5, fig.height=3}
plotReachAftereffects(selectPerformance=selectPerformance)
```

@@

The figure looked like there might be some difference between the two training types, mainly in where the generalization curves peak. There is no main effect of training type, so the overall magnitude of reach aftereffects is not different. But there is an interaction between training and target (p=.038), so that perhaps the curves peak at different locations.

So we now investigate the peaks the confidence intervals for the peaks of both:

@@

The 95% confidence interval for the classically trained group includes the 45 degree target (where the visual feedback was during training) but not the 75 degree target, whereas the 95% confidence interval for the exposure training group includes the 75 degree target (where the hand really was during training) but not the 45 degree target. Something might be going on -- but the two confidence intervals are not different from each other, which is not in line with the LME results. We can't conclude much, except that we don't have strong evidence for different generalization curves.

1. Rotated exposure training evokes substantial shifts in no-cursor reach direction across the workspace,
2. these changes persist over time, and
3. these changes are comparable to the ones we found after classic training, and
4. there is some generalization curve across the workspace, but
5. we have no evidence that this is different from reach aftereffect generalization after classic training.

@@

1. Exposure training leads to shifts in localization, that are not very different between active and passive localization. Since both active and passive movements access recalibrated proprioception, while active localization also accesses updated predictions, this indicates that hand localization is makes very good use of proprioception.

@@

2. Exposure training leads to robust reach aftereffects that are similar to those after classic training, although they seem to have a different generalization curve. This means that evoking propceptive recalibration with visual-proprioceptive discrepancies leads to motor changes. In other words: recalibrated proprioception informs motor control.

    1. The generalization curves of reach aftereffects look somewhat different for both groups, which can be explained by the effect of updated predictions that factors into the curve for the classic training group, but not the exposure training group. However, it's not a large difference, and the analyses are ambiguous. Perhaps this is another cue that recalibrated proprioception dominates even reach aftereffects? Either way, we find no (statistically reliable) evidence that the updated predictions matter much for the generalization curves, which can have several causes.

```{r create_svgs, cache=FALSE, include=FALSE}
# This chunk creates SVG files but shouldn't show up in the PDF / HTML 
plotReachAftereffects(generateSVG=TRUE,
                      selectPerformance=selectPerformance)
plotLocalization(classicOnline=FALSE,
                 generateSVG=TRUE, 
                 selectPerformance=selectPerformance)
correlations <- correlateNoCursorsLocalization(generateSVG=TRUE, 
                                               selectPerformance=selectPerformance)
```

>>>>

These are the planned figures. Figures one and two will not be included here as they are not generated from data but illustrate the setup, tasks and experimental design.

1. Setup & Task (goal: visualize the experiment)
    + A: Setup (side view)
    + B: Rotated training (top view)
    + C: No-cursor reaches (top view)
    + D: Localization (top view)
2. Experimental Design
    + A: Task order (instead of the table in the previous paper)
    + B: Main localization measures
3. Localization
    + A: Exposure, active and passive (goal: small/no difference between active and passive)
    + B: Classic, active and passive (goal: larger difference between active and passive, different generalization?)
4. Reach Aftereffects (goal: show motor changes are there, are robust and compare to classic)
    + A: Exposure reach aftereffects (initial & later)
    + B: Classic & Exposure reach aftereffects (initial only? or all?)
5. Generalization Maxima
    + A: Maxima of active localization after classic and exposure training
    + B: Maxima of reach aftereffects after classic and exposure training

@@

6. Correlation
    + A: Exposure, active and passive localization predicting reach aftereffects
    + B: Classic, active and passive localization predicting reach aftereffects
7. Individual localization responses

@@

The scripts doing the statistics and creating figures are in a few separate R files. They allow many options, but here we will only see / do the ones that we think should go to the manuscript.

@@

```{r require_lme_packages}
required.packages = c('lme4', 'lmerTest')
installRequire.Packages(required.packages, installmissing=FALSE)
if (!all(required.packages %in% installed.packages())) {
  warning('packages required for analysis are not installed, later chunks will fail\n')
}
```

@@

In the manuscript we'll first show that there indeed are localization shifts after exposure training (with the no-cursor data) and then compare that with the classic. Second are the no-cursor reaches (and the reach aftereffects). Then we'll look at generalization patterns of localization shifts and reach aftereffects after both classic and exposure training.

A further topic we only explore in this notebook is if localization shifts can predict no-cursor changes:

1. Analyses (some general remarks about analyses)
2. Participants (participant demographics)
3. Localization
4. Reach aftereffects
5. Generalization Maxima
6. Correlations

@@

Since in the localization tasks there is missing data at some points (usually at the extremes) we can't use ANOVAs directly on the data. Imputation is going to be very hard because the missing data is systematically missing, and there is no (good) model for the localization data (for reach aftereffects it might work). An alternative to imputation is to either remove participants or removed measurement locations, but that would mean deleting 7/21 participants or 4/7 locations, so at least 1/3 of the data, which would mean a big loss in power. As the data is systematically missing, this might also skew results away from what would be observed in the general population.

Hence, we are using Linear Mixed Effects models as an alternative, but the output of those models are not readily interpretable by everybody. Initially we decided to use the $\chi^2$ approach (from the `Anova` function from the `car` package). However, a paper from Steven G. Luke (2017, _Evaluating significance in linear mixed-effects models in R_) suggests using Satterthwaite approximation. So all functions running an LME now take an argument that tells them to use the Satterthwaite method, but it can be set to chi-squared as well. It turns out that it doesn't matter much, so it's a good sanity check. For consistency across all the analyses, we set the value of that argument here:

@@

```{r requireChiSquared_packages}
if (LMEmethod=='chi-squared') {
  required.packages = c('nlme', 'car')
  installRequire.Packages(required.packages, installmissing=FALSE)
  # we check if that worked and otherwise
  if (!all(required.packages %in% installed.packages())) {
    LMEmethod <- 'Satterthwaite'
  }
}
```

If the packages were not installed (we won't force you), we do switch back to the Satterthwaite method so that we don't generate errors instead of statistical results:

LME analyses have fixed effects, usually the effects we are interested in, as well as random effects, which are usually not the effects we are interested in. Fixed effects are comparable to 'factors' in ANOVA's, and including random effects is conceptually (but not mathematically) comparable to making it a repeated measures design, or including covariates. We'll spell out the fixed and random effects included in each analyses right before showing the results, so that those results can easily be put in the proper context by the reader, and hence not all details are mentioned in the methods section.

### Generalization Maxima

Since participants trained with a single target, we could look at generalization curves. These appear to be shifted more counter clockwise after exposure training compared to classic, so we want to quantify this. But, for most of the curves, and especially the ones that seems to be shifted, we don't have data on the full curve on one side of the peak, which limits the interpretability of the analysis somewhat. We can still bootstrap the __peak__ of Guassian functions fit to the data across many samples of participants for the exposure group and for the delayed localization after classic training, but not for the online localization after classic training (the shifts are even larger there...). This data is presented in the paper, where we only look at the peak of the localization -- not the width of the generalization curve.

The range of the data doesn't include the peak in some participants, which makes it hard to fit a normal function to every individual participant's data. In one alternative approach, we assumed that the minimal change in the data, occurs far beyond the boundaries of the measured workspace, and converted the data to a cumulative distribution, assuming that at some point it would revert back to the base level we see on the other side of the peak. We can scale back to a range from 0 to 1, to allow fitting a _cumulative_ normal function where the 50% point corresponds to the peak of the generalization curve. This method is more stable, and seems to works fairly well on the datasets where the more common fits also work (but we didn't test online localization from the classic experiment). Code for this approach is available (functions `psychometricGeneralization()`, `pnormMSE()` and `getCDFpeaks()` at the end of the file: `shared.R`), but *these analysis are not presented* in the paper or notebook, but the code is left here for the interested reader.

@@

Here are some simple descriptives of participants, as well as the selection of participants based on how they performed in detecting the very brief blink of the cursor (2 frames ~33 ms) during the exposure training. This ensures that the participants paid attention to the cursor during training, which is probably required for the cross-sensory error signal to have the desired effect of proprioceptive recalibration.

@@

We'll also set a variable that is to be passed to all functions to exclude the underperformers:

@@

This is highly similar between conditions and training paradigms, and at an acceptable level.

@@

These high percentages may indicate there are some additional no-cursor reaches that could be considered outliers and romeving them would not constitute a big loss of data. However, as it is, the variance in the dataset appears to mostly reflect inidividual differences and not noise, so we kept the data as is.

@@

@@

Here is a plot, probably Figure 4:

```{r fig4_reachaftereffect, fig.width=7.5, fig.height=3}
plotReachAftereffects(selectPerformance=selectPerformance)
```

@@

So there is no effect of training (overall reach aftereffect magnitudes are not very different between classic and exposure training), but there is an effect of target (there is a generalization curve, as we can clearly see in the data) and there is an interaction between training and target (the generalization curves are different between classic and exposure training). This last bit is very similar to the pattern observed in localization, but we'll get back to that later.

The figure looked like there might be some difference between the two training types, mainly in where the generalization curves peak. There is no main effect of training type, so the overall magnitude of reach aftereffects is not different.

However, the main take-away message from this analysis is that the motor changes measured with changes in open-loop reach direction are of similar magnitude in both types of training. That is, passive exposure training (proprioceptive recalibration) drives motor changes.

1. Rotated exposure training evokes substantial shifts in no-cursor reach direction across the workspace,
2. these changes persist over time, and
3. these changes are comparable to the ones we found after classic training, and
4. there is some generalization curve across the workspace, but

# Generalization Maxima

Here we'll look at the differences in generalization suggested by both the LME on localization shift and on reach aftereffects. In particular, we estimate the peak of each curve by bootstrapping participants and fitting a Gaussian (normal) function a 1000 times. This disregards other possible differences in generalization curves: the width of the curve, the peak amplitude or the offset. There may be effects there, but we are most interested in the location of the maxima.

First, a figure:

```{r fig5_generalization_maxima, fig.width=7.5, fig.height=3}
plotGeneralization(generateSVG=FALSE,
                   selectPerformance=selectPerformance)
```

## Localization

There was an interaction between group and hand angle on localization shifts in the LME (far) above, which suggests different generalization curves in the two groups. Here are the confidence intervals for the peaks of the generalization curves for _active_ localization:

The median values are a little over 13 degrees apart, and median for classic falls just outside the exposure confidence interval. This could explain the interaction between group and hand angle on localization shifts, but the other way around it's not so clear.

## Reach aftereffects

Now we look at the confidence intervals for the maxima of the generalization curves of reach aftereffects after both exposure training and after classic training:

@@

The 95% confidence interval for the classically trained group includes the 45 degree target (where the visual feedback was during training) but not the 75 degree target, whereas the 95% confidence interval for the exposure training group includes the 75 degree target (where the hand really was during training) but not the 45 degree target. Something might be going on -- but the two confidence intervals are not different from each other, which is not in line with the LME results. We can't conclude much, except that we don't know if the differences in generalization curves comes from a difference in the location of the maxima, or perhaps the width of the curves, or both.

@@

1. **Exposure training leads to shifts in localization**, that are _indistinguishable_ for active and passive localization. Since both active and passive movements access recalibrated proprioception, while active localization also accesses updated predictions, this indicates that: _hand localization relies on (recalibrated) proprioception_.

@@

2. **Exposure training leads to robust reach aftereffects** that are _indistinguishable_ in magnitude to those after classic training (although they seem to have a different generalization curve). This means that evoking propceptive recalibration with visual-proprioceptive discrepancies leads to motor changes. In other words: _recalibrated proprioception informs motor control_.

    1. The generalization curves of reach aftereffects look somewhat different for both groups, which can be explained by the effect of updated predictions that factors into the curve for the classic training group, but not the exposure training group. However, it's not a large difference, and the analyses are ambiguous. Either way, we find no (statistically reliable) evidence that the updated predictions matter much for the generalization curves, which can have several causes all outside the scope of this paper.

```{r create_svgs, cache=FALSE, include=FALSE}
# This chunk creates SVG files but shouldn't show up in the PDF / HTML 
plotLocalization(classicOnline=FALSE,
                 generateSVG=TRUE, 
                 selectPerformance=selectPerformance)
plotReachAftereffects(generateSVG=TRUE,
                      selectPerformance=selectPerformance)
plotGeneralization(generateSVG=TRUE,
                   selectPerformance=selectPerformance)
correlations <- correlateNoCursorsLocalization(generateSVG=TRUE, 
                                               selectPerformance=selectPerformance)
```
